---
title: "Generalized linear mixed-effects modeling of the composite face task"
author: "[Haiyang Jin](https://haiyangjin.github.io/)"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_float: true
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 2000)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, include = TRUE,
                      cache = FALSE, tidy = FALSE, size = "small")
```

# Preparation
```{r setup and load the related libraries, include=FALSE}
library(tidyverse)
library(hypr)
library(MASS)
library(lme4)
library(lmerTest)
library(optimx)
library(emmeans)

# set the contrast coding
options(contrasts = rep("contr.sdif", 2))
```

## Prepare data
```{r read the data file}
# backup codes
# df_tmp <- read_csv(file.path("data", "example.csv")) %>%
#   transmute(Participant = as_factor(Participant),
#             Condition = if_else(substr(Experiment,1,3)=="104", "CFS", "monocular"),
#             Congruency = if_else(Congruency=="C", "congruent", "incongruent"),
#             Alignment = if_else(Alignment=="A", "aligned", "misaligned"),
#             answer = if_else(SameDifferent=="S", "same", "different"),
#             resp = if_else(thisResponse=="S", 1, 0),
#             isCorrect = isCorrect,
#             RT = round(reactionTime*1000+200)) %>%
#   mutate(Participant = sprintf("Subj%02d", as.numeric(Participant)))
# write_csv(df_tmp, file = file.path("data", "example_data.csv"))

folder_out <- "output"

df_raw <-  read_csv(file.path("data", "example_data.csv")) %>% 
  mutate(Condition = if_else(Condition=="monocular", "mon", Condition),
         Congruency = if_else(Congruency=="congruent", "con", "inc"),
         Alignment = if_else(Alignment=="aligned", "ali", "mis"),
         answer = if_else(answer=="same", "sam", "dif"),
         Condition = factor(Condition, levels=c("CFS", "mon")),
         Congruency = factor(Congruency, levels=c("inc", "con")))
head(df_raw)
```

## Brief introduction
The data was a subset of [one previous study](https://psyarxiv.com/j8g6z/) where data was analyzed with repeated-measures ANOVA. The independent variables are:
  
+ Condition: monocular (**O**) vs. CFS (**F**)  
+ Congruency: congruent (**C**) vs. incongruent (**I**)  
+ Alignment: aligned (**A**) vs. misaligned (**M**)  
+ answer: same (**S**) vs. different (**D**) (used as `signal` and `noise` in Signal Detection Theory)  

The dependent variables are behavioral responses and response times. Behavioral responses will be analyzed with signal detection theory (i.e., generalized linear mixed-effects models with `probit` link). Response times will be analyzed with lognormal transformation.

The main research questions are: 
(1) whether there is composite face effect in the `CFS` condition;
(2) whether there is composite face effect in the `monocular` condition;
(3) whether the composite face effect is larger in the `monocular` relative to the `CFS` condition. 

As discussed [earlier](https://psyarxiv.com/yhmzg/), we may claim observing the composite face effect in a particular condition (e.g., `CFS` or `monocular`) only when (1) the performance for aligned faces is better in the congruent relative to the incongruent condition (i.e., `congruent_aligned` > `incongruent_aligned`; `CA` > `IA`) and (2) the increased performance in congruent relative to incongruent condition is larger for aligned compared to misaligned faces (i.e., (`congruent_aligned` - `incongruent_aligned`) > (`congruent_misaligned` - `incongruent_misaligned`); (`CA` - `IA`) > (`CM` - `IM`)). 

Moreover, for the third question, we also need to examine whether the composite effect in the `monocular` condition is larger than that in the `CFS` condition (i.e., [(`monocular_congruent_aligned` - `monocular_incongruent_aligned`) - (`monocular_congruent_misaligned` - `monocular_incongruent_misaligned`)] > [(`CFS_congruent_aligned` - `CFS_incongruent_aligned`) - (`CFS_congruent_misaligned` - `CFS_incongruent_misaligned`)]; [(`OCA` - `OIA`) - (`OCM` - `OIM`)] > (`FCA` - `FIA`) - (`FCM` - `FIM`)]). 


# GLMM with probit link
sensitivity d'

## Priori contrasts

```{r}
# levels for the new factor (see below)
 level_resp <- c("OCAS", "OCAD", "OIAS", "OIAD", "OCMS", "OCMD", "OIMS", "OIMD", "FCAS", "FCAD", "FIAS", "FIAD", "FCMS", "FCMD", "FIMS", "FIMD")

hypr_resp <- hypr(OCAS - OCAD - OIAS + OIAD ~ 0,
                  OCAS - OCAD - OIAS + OIAD - OCMS + OCMD + OIMS - OIMD ~ 0,
                  FCAS - FCAD - FIAS + FIAD ~ 0,
                  FCAS - FCAD - FIAS + FIAD - FCMS + FCMD + FIMS - FIMD ~ 0,
                  OCAS - OCAD - OIAS + OIAD - OCMS + OCMD + OIMS - OIMD - (FCAS - FCAD - FIAS + FIAD - FCMS + FCMD + FIMS - FIMD) ~ 0,
                  levels =  level_resp
)
hypr_resp
```


```{r}
df_resp <- df_raw %>% 
  mutate(I_ = paste0(if_else(Condition == "monocular", "O", "F"), 
                     if_else(Congruency == "congruent", "C", "I"),
                     if_else(Alignment == "alignment", "A", "M"),
                     if_else(answer == "same", "S", "D")),
         I_ = factor(I_, levels= level_resp))
```

```{r}
contrasts(df_resp$I_) <- contr.hypothesis(hypr_resp)
```


# Lognormal

```{r}
# levels for the new factor (see below)
levels_rt <- c("OCA", "OIA", "OCM", "OIM", "FCA", "FIA", "FCM", "FIM")

# make the level names shorter
df_rt <- df_raw %>% 
  mutate(I_ = paste0(if_else(Condition == "mon", "O", "F"), 
                     if_else(Congruency == "con", "C", "I"),
                     if_else(Alignment == "ali", "A", "M")),
         I_ = factor(I_, levels= levels_rt)) %>% 
  filter(isCorrect == 1)
```


# try

```{r}
model.matrix(~ Congruency, df_raw, contrasts.arg = list(Congruency="contr.sum")) %>% 
  as.data.frame() %>% 
  distinct()
```


```{r}
df_tmp <- df_raw %>% 
  filter(Condition == "monocular")

aov_tmp <- aov_4(isCorrect ~ Congruency * Alignment * answer + 
                   (Congruency * Alignment * answer | Participant), 
                 data = df_tmp, 
                 fun_aggregate = mean)

```


```{r}
emm_tmp <- emmeans(aov_tmp, ~ Alignment | Congruency + answer)

contrast(emm_tmp, "pairwise")[1:4]
```


```{r}
model.matrix(~ Condition / Congruency * Alignment, df_rt) %>% 
  as_tibble() %>% 
  distinct() %>% 
  as.matrix() %>% 
  ginv2()
```

```{r}
options(contrasts=rep("contr.treatment", 2))
model.matrix(~ Condition / Congruency * Alignment, df_rt) %>% 
  as_tibble() %>% 
  distinct() %>% 
  as.matrix() %>% 
  ginv2()
```


```{r}


hypr_rt1 <- hypr((OCA + OIA + OCM + OIM + FCA + FIA + FCM + FIM)/8 ~ 0, # intercept
                OCA - OIA ~ 0,
                OCA - OIA - OCM + OIM ~ 0,
                FCA - FIA ~ 0,
                # FCA - FIA - FCM + FIM ~ 0,
                OCA - OIA - OCM + OIM - (FCA - FIA - FCM + FIM) ~ 0,
                
                levels = levels_rt)
hypr_rt
```

```{r}
cor(contr.hypothesis(hypr_resp))
hypr_rt1
```



```{r}
contrasts(df_rt$I_) <- contr.hypothesis(hypr_rt)
```



**Question**: According to my research question, I need to test the five effects discussed earlier (i.e., the first five columns in `hypr_resp`). How should I decide the other (10) effects tested in the model?

```{r}
# apply the custom contrast to raw data
contrasts(df_raw$I_) <- con_resp

# create variables for each parameter
df_resp <- as_tibble(model.matrix(~ 1 + I_, df_raw)) %>% 
  cbind(df_raw, .)
```



for response times
```{r only keep correct trials (but with both same and different trials) }
df_rt <- df_raw %>% 
  filter(isCorrect == 1)

```




```{r  rt zcp}
file_rt_zcp <- file.path(folder_lmm, "rt_lmm_zcp.RData")

# fit the zcp1 model
if (!file.exists(file_rt_zcp)) {
  glmm_rt_zcp <- lmer(
    # log(rt) ~ Cue * Congruency * Alignment +  
    #   (Cue_C + Con_C + Ali_C + 
    #      Cue_Con + Cue_Ali + Con_Ali + 
    #      Cue_Con_Ali || Participant),
    data = df_lmm,
    control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                          optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )
  # save(glmm_rt_zcp, file = file_rt_zcp)
} else {
  load(file_rt_zcp)
}

print(summary(glmm_rt_zcp), corr = FALSE)
```


# Session information {.unlisted .unnumbered}
```{r}
sessionInfo()
```


