---
title: "Generalized linear mixed-effects modeling of the composite face task"
author: "[Haiyang Jin](https://haiyangjin.github.io/)"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    code_folding: hide
    number_sections: true
    toc: true
    toc_float: true
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 2000)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE,
                      cache = FALSE, tidy = FALSE, size = "small")
```

# Preparation
```{r setup and load the related libraries, include=FALSE}
library(tidyverse)
library(hypr)
library(MASS)
library(lme4)
library(lmerTest)
library(optimx)
library(emmeans)
library(arrow)

# set the contrast coding
options(contrasts = rep("contr.sdif", 2))
```

## Prepare data
```{r read the data file}
# backup codes
# df_tmp <- read_csv(file.path("data", "example.csv")) %>%
#   transmute(Participant = as_factor(Participant),
#             Condition = if_else(substr(Experiment,1,3)=="104", "CFS", "monocular"),
#             Congruency = if_else(Congruency=="C", "congruent", "incongruent"),
#             Alignment = if_else(Alignment=="A", "aligned", "misaligned"),
#             answer = if_else(SameDifferent=="S", "same", "different"),
#             resp = if_else(thisResponse=="S", 1, 0),
#             isCorrect = isCorrect,
#             RT = round(reactionTime*1000+200)) %>%
#   mutate(Participant = sprintf("Subj%02d", as.numeric(Participant)))
# write_csv(df_tmp, file = file.path("data", "example_data.csv"))

folder_out <- "cache"

df_raw <-  read_csv(file.path("data", "example_data.csv")) %>% 
  mutate(Condition = if_else(Condition=="monocular", "mon", Condition),
         Congruency = if_else(Congruency=="congruent", "con", "inc"),
         Alignment = if_else(Alignment=="aligned", "ali", "mis"),
         answer = if_else(answer=="same", "sam", "dif"),
         Condition = factor(Condition, levels=c("CFS", "mon")),
         Congruency = factor(Congruency, levels=c("inc", "con")),
         Alignment = factor(Alignment, levels=c("mis", "ali")))
head(df_raw)
```

## Brief introduction
The data were a subset of [one previous study](https://psyarxiv.com/j8g6z/){target="_blank"} where data were analyzed with repeated-measures ANOVA. The independent variables are:
  
+ Condition: monocular (**O**) vs. CFS (**F**)  
+ Congruency: congruent (**C**) vs. incongruent (**I**)  
+ Alignment: aligned (**A**) vs. misaligned (**M**)  
+ answer: same (**S**) vs. different (**D**) (used as `signal` and `noise` in Signal Detection Theory)  

The dependent variables are behavioral responses and response times. Behavioral responses are analyzed with signal detection theory (i.e., generalized linear mixed-effects models with `probit` link). Response times are analyzed with log-normal transformation.

The main research questions are: 
(1) whether there is composite face effect in the `CFS` condition;
(2) whether there is composite face effect in the `monocular` condition;
(3) whether the composite face effect is larger in the `monocular` than the `CFS` condition. 

As discussed [earlier](https://psyarxiv.com/yhmzg/){target="_blank"}, we may claim observing the composite face effect in a particular condition (e.g., `CFS` or `monocular`) only when (1) the performance for aligned faces is better in the congruent relative to the incongruent condition (i.e., `congruent_aligned` > `incongruent_aligned`) and (2) the increased performance in congruent relative to incongruent condition is larger for aligned compared to misaligned faces (i.e., (`congruent_aligned` - `incongruent_aligned`) > (`congruent_misaligned` - `incongruent_misaligned`)). 

Moreover, for the third question, we also need to examine whether the composite effect in the `monocular` condition is larger than that in the `CFS` condition (i.e., [(`monocular_congruent_aligned` - `monocular_incongruent_aligned`) - (`monocular_congruent_misaligned` - `monocular_incongruent_misaligned`)] > [(`CFS_congruent_aligned` - `CFS_incongruent_aligned`) - (`CFS_congruent_misaligned` - `CFS_incongruent_misaligned`)]). 

# Sensitivity d': GLMM with probit link
This section analyzes the behavioral responses with signal detection theory (with `probit` link): `same` and `different` trials are used as `signal` and `noise` to calculate the sensitivity d'.

## lmm for simple effects of Congruency 
It is also testing the interaction between Congruency and Answer.
```{r}
df_resp_2inter <- model.matrix(~ Condition * Alignment / Congruency * answer, df_raw) %>% 
  as_tibble() %>% 
  cbind(df_raw, .)
```
`df_resp_2inter` was created to facilitate simplifying random effects later.

```{r}
# save data as arrow
write_arrow(df_resp_2inter, file.path("data", "example_nested.arrow"))
```

### The maximal model
```{r resp1 max d }
# file_resp1_max <- file.path(folder_out, "Resp1_lmm_max.RData")
# 
# # fit the max model
# if (!file.exists(file_resp1_max)) {
#   glmm_resp1_max <- glmer(resp ~ Condition * Alignment / Congruency * answer +
#                            (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` +
#                               `Conditionmon-CFS:Alignmentali-mis` +
#                               `Conditionmon-CFS:answersam-dif` +
#                               `Alignmentali-mis:answersam-dif` +
#                               `ConditionCFS:Alignmentali:Congruencycon-inc` +
#                               `Conditionmon:Alignmentali:Congruencycon-inc` +
#                               `ConditionCFS:Alignmentmis:Congruencycon-inc` +
#                               `Conditionmon:Alignmentmis:Congruencycon-inc` +
#                               `Conditionmon-CFS:Alignmentali-mis:answersam-dif` +
#                               `ConditionCFS:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
#                               `Conditionmon:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
#                               `ConditionCFS:Alignmentmis:Congruencycon-inc:answersam-dif` +
#                               `Conditionmon:Alignmentmis:Congruencycon-inc:answersam-dif`
#                               | Participant),
#     family = binomial(link = "probit"),
#     data = df_resp_2inter,
#     control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
#                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
#   )
# 
#   save(glmm_resp1_max, file = file_resp1_max)
# } else {
#   load(file_resp1_max)
# }
# 
# print(summary(glmm_resp1_max), corr = FALSE)
```

### The zero-correlation-parameter model
```{r resp1 zcp d }
file_resp1_zcp <- file.path(folder_out, "Resp1_lmm_zcp.RData")

# fit the zcp model
if (!file.exists(file_resp1_zcp)) {
  glmm_resp1_zcp <- glmer(resp ~ Condition * Alignment / Congruency * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` + 
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Alignmentali:Congruencycon-inc` +
                              `Conditionmon:Alignmentali:Congruencycon-inc` +
                              `ConditionCFS:Alignmentmis:Congruencycon-inc` +
                              `Conditionmon:Alignmentmis:Congruencycon-inc` +
                              `Conditionmon-CFS:Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              `Conditionmon:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              `ConditionCFS:Alignmentmis:Congruencycon-inc:answersam-dif` +
                              `Conditionmon:Alignmentmis:Congruencycon-inc:answersam-dif` 
                              || Participant),
    family = binomial(link = "probit"),
    data = df_resp_2inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp1_zcp, file = file_resp1_zcp)
} else {
  load(file_resp1_zcp)
}

print(summary(glmm_resp1_zcp), corr = FALSE)
```

### The reduced model
```{r PCA analysis for resp1 zcp lmm }
summary(rePCA(glmm_resp1_zcp))
```

Random effect that explained less than 0.1% variances were removed from the zcp model, leading to the reduced model.

```{r resp1 rdc d }
file_resp1_rdc <- file.path(folder_out, "Resp1_lmm_rdc.RData")

# fit the rdc model
if (!file.exists(file_resp1_rdc)) {
  glmm_resp1_rdc <- glmer(resp ~ Condition * Alignment / Congruency * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` + 
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Alignmentali:Congruencycon-inc` +
                              # `Conditionmon:Alignmentali:Congruencycon-inc` +
                              `ConditionCFS:Alignmentmis:Congruencycon-inc` # +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc` +
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` +
                              # `ConditionCFS:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `Conditionmon:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `ConditionCFS:Alignmentmis:Congruencycon-inc:answersam-dif` +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc:answersam-dif` 
                              || Participant),
    family = binomial(link = "probit"),
    data = df_resp_2inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp1_rdc, file = file_resp1_rdc)
} else {
  load(file_resp1_rdc)
}

print(summary(glmm_resp1_rdc), corr = FALSE)
```

### The extended model
```{r resp1 etd d }
file_resp1_etd <- file.path(folder_out, "Resp1_lmm_etd.RData")

# fit the etd model
if (!file.exists(file_resp1_etd)) {
  glmm_resp1_etd <- glmer(resp ~ Condition * Alignment / Congruency * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` + 
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Alignmentali:Congruencycon-inc` +
                              # `Conditionmon:Alignmentali:Congruencycon-inc` +
                              `ConditionCFS:Alignmentmis:Congruencycon-inc` # +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc` +
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` +
                              # `ConditionCFS:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `Conditionmon:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `ConditionCFS:Alignmentmis:Congruencycon-inc:answersam-dif` +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc:answersam-dif` 
                              | Participant),
    family = binomial(link = "probit"),
    data = df_resp_2inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp1_etd, file = file_resp1_etd)
} else {
  load(file_resp1_etd)
}

print(summary(glmm_resp1_etd), corr = FALSE)
```

```{r}
summary(rePCA(glmm_resp1_etd))
```

```{r resp1 etd1 d }
file_resp1_etd1 <- file.path(folder_out, "Resp1_lmm_etd1.RData")

# fit the etd1 model
if (!file.exists(file_resp1_etd1)) {
  glmm_resp1_etd1 <- glmer(resp ~ Condition * Alignment / Congruency * answer +
                           (`Conditionmon-CFS` + `answersam-dif` + # `Alignmentali-mis` + 
                              # `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` # +
                              # `Alignmentali-mis:answersam-dif` +
                              # `ConditionCFS:Alignmentali:Congruencycon-inc` +
                              # `Conditionmon:Alignmentali:Congruencycon-inc` +
                              # `ConditionCFS:Alignmentmis:Congruencycon-inc` # +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc` +
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` +
                              # `ConditionCFS:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `Conditionmon:Alignmentali:Congruencycon-inc:answersam-dif` + # interest
                              # `ConditionCFS:Alignmentmis:Congruencycon-inc:answersam-dif` +
                              # `Conditionmon:Alignmentmis:Congruencycon-inc:answersam-dif` 
                              | Participant),
    family = binomial(link = "probit"),
    data = df_resp_2inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp1_etd1, file = file_resp1_etd1)
} else {
  load(file_resp1_etd1)
}

print(summary(glmm_resp1_etd1), corr = FALSE)
```


### The optimal model
```{r comapre etd and rdc resp1 d}
# compare the extended and reduced model
anova(glmm_resp1_etd1, glmm_resp1_rdc, refit = FALSE)
```

```{r the optimal model resp1 d}
glmm_resp1_opt <- glmm_resp1_etd1

print(summary(glmm_resp1_opt), corr = FALSE)
```
The above results (the third last row) showed that the sensitivity d' is larger for congruent relative to incongruent trials for aligned faces in the monocular condition. 

## lmm for simple effects of Congruency 
It is also testing the interaction between Congruency and Answer.
```{r}
df_resp_3inter <- model.matrix(~ Condition / Congruency * Alignment * answer, df_raw) %>% 
  as_tibble() %>% 
  cbind(df_raw, .)
```
`df_resp_3inter` was created to facilitate simplifying random effects later.

### The maximal model
```{r resp2 max d }
# file_resp2_max <- file.path(folder_out, "Resp2_lmm_max.RData")
# 
# # fit the max model
# if (!file.exists(file_resp2_max)) {
#   glmm_resp2_max <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
#                            (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` +
#                               `ConditionCFS:Congruencycon-inc` +
#                               `Conditionmon:Congruencycon-inc` +
#                               `Conditionmon-CFS:Alignmentali-mis` +
#                               `Conditionmon-CFS:answersam-dif` +
#                               `Alignmentali-mis:answersam-dif` +
#                               `ConditionCFS:Congruencycon-inc:Alignmentali-mis` +
#                               `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
#                               `ConditionCFS:Congruencycon-inc:answersam-dif` +
#                               `Conditionmon:Congruencycon-inc:answersam-dif` + 
#                               `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
#                               `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
#                               `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
#                               | Participant),
#     family = binomial(link = "probit"),
#     data = df_resp_3inter,
#     control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
#                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
#   )
# 
#   save(glmm_resp2_max, file = file_resp2_max)
# } else {
#   load(file_resp2_max)
# }
# 
# print(summary(glmm_resp2_max), corr = FALSE)
```

### The zero-correlation-parameter model
```{r resp2 zcp d }
file_resp2_zcp <- file.path(folder_out, "Resp2_lmm_zcp.RData")

# fit the zcp model
if (!file.exists(file_resp2_zcp)) {
  glmm_resp2_zcp <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` +
                              `ConditionCFS:Congruencycon-inc` +
                              `Conditionmon:Congruencycon-inc` +
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis` +
                              `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
                              `ConditionCFS:Congruencycon-inc:answersam-dif` +
                              `Conditionmon:Congruencycon-inc:answersam-dif` + 
                              `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
                              `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
                              || Participant),
    family = binomial(link = "probit"),
    data = df_resp_3inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp2_zcp, file = file_resp2_zcp)
} else {
  load(file_resp2_zcp)
}

print(summary(glmm_resp2_zcp), corr = FALSE)
```

### The reduced model
```{r PCA analysis for resp2 zcp lmm}
summary(rePCA(glmm_resp2_zcp))
```

Random effect that explained less than 0.1% variances were removed from the zcp model, leading to the reduced model.

```{r resp2 rdc d }
file_resp2_rdc <- file.path(folder_out, "Resp2_lmm_rdc.RData")

# fit the rdc model
if (!file.exists(file_resp2_rdc)) {
  glmm_resp2_rdc <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` +
                              `ConditionCFS:Congruencycon-inc` +
                              `Conditionmon:Congruencycon-inc` +
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis` +
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
                              `ConditionCFS:Congruencycon-inc:answersam-dif` # +
                              # `Conditionmon:Congruencycon-inc:answersam-dif` + 
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
                              # `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
                              || Participant),
    family = binomial(link = "probit"),
    data = df_resp_3inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp2_rdc, file = file_resp2_rdc)
} else {
  load(file_resp2_rdc)
}

print(summary(glmm_resp2_rdc), corr = FALSE)
```

### The extended model
```{r resp2 etd d }
file_resp2_etd <- file.path(folder_out, "Resp2_lmm_etd.RData")

# fit the etd model
if (!file.exists(file_resp2_etd)) {
  glmm_resp2_etd <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
                           (`Conditionmon-CFS` + `Alignmentali-mis` + `answersam-dif` +
                              `ConditionCFS:Congruencycon-inc` +
                              `Conditionmon:Congruencycon-inc` +
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis` +
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
                              `ConditionCFS:Congruencycon-inc:answersam-dif` # +
                              # `Conditionmon:Congruencycon-inc:answersam-dif` + 
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
                              # `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
                              | Participant),
    family = binomial(link = "probit"),
    data = df_resp_3inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp2_etd, file = file_resp2_etd)
} else {
  load(file_resp2_etd)
}

print(summary(glmm_resp2_etd), corr = FALSE)
```

```{r}
summary(rePCA(glmm_resp2_etd))
```

```{r resp2 etd1 d }
file_resp2_etd1 <- file.path(folder_out, "Resp2_lmm_etd1.RData")

# fit the etd1 model
if (!file.exists(file_resp2_etd1)) {
  glmm_resp2_etd1 <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
                           (`Conditionmon-CFS` + `answersam-dif` + # `Alignmentali-mis` + 
                              # `ConditionCFS:Congruencycon-inc` +
                              # `Conditionmon:Congruencycon-inc` +
                              `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              # `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis` # +
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
                              # `ConditionCFS:Congruencycon-inc:answersam-dif` # +
                              # `Conditionmon:Congruencycon-inc:answersam-dif` + 
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
                              # `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
                              | Participant),
    family = binomial(link = "probit"),
    data = df_resp_3inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp2_etd1, file = file_resp2_etd1)
} else {
  load(file_resp2_etd1)
}

print(summary(glmm_resp2_etd1), corr = FALSE)
```
```{r}
summary(rePCA(glmm_resp2_etd1))
```

```{r resp2 etd2 d }
file_resp2_etd2 <- file.path(folder_out, "Resp2_lmm_etd2.RData")

# fit the etd2 model
if (!file.exists(file_resp2_etd2)) {
  glmm_resp2_etd2 <- glmer(resp ~ Condition / Congruency * Alignment  * answer +
                           (`answersam-dif` + # `Alignmentali-mis` +`Conditionmon-CFS` + 
                              # `ConditionCFS:Congruencycon-inc` +
                              # `Conditionmon:Congruencycon-inc` +
                              # `Conditionmon-CFS:Alignmentali-mis` +
                              `Conditionmon-CFS:answersam-dif` +
                              # `Alignmentali-mis:answersam-dif` +
                              `ConditionCFS:Congruencycon-inc:Alignmentali-mis` # +
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis` +
                              # `ConditionCFS:Congruencycon-inc:answersam-dif` # +
                              # `Conditionmon:Congruencycon-inc:answersam-dif` + 
                              # `Conditionmon-CFS:Alignmentali-mis:answersam-dif` + 
                              # `ConditionCFS:Congruencycon-inc:Alignmentali-mis:answersam-dif` + # interest
                              # `Conditionmon:Congruencycon-inc:Alignmentali-mis:answersam-dif` # interest
                              | Participant),
    family = binomial(link = "probit"),
    data = df_resp_3inter,
    control = glmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                           optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(glmm_resp2_etd2, file = file_resp2_etd2)
} else {
  load(file_resp2_etd2)
}

print(summary(glmm_resp2_etd2), corr = FALSE)
```

Though `glmm_resp2_etd2` does not converge, the `glmm_resp2_rdc` already explains the data better (than `glmm_resp2_etd2`).

### The optimal model
```{r comapre etd and rdc resp2 d}
# compare the extended and reduced model
anova(glmm_resp2_etd2, glmm_resp2_rdc, refit = FALSE)
```


```{r the optimal model resp2 d}
glmm_resp2_opt <- glmm_resp2_rdc

print(summary(glmm_resp2_opt), corr = FALSE)
```
The above results (the last row) showed that in the monocular condition, sensitivity d' difference between congruent and incongruent trials was larger for aligned relative to misaligned faces.  

## lmm for comparing composite effects in CFS and monocular conditions
This analysis will not be repeated; it is the same as that used in `lmm_CF_emmeans.Rmd` file (around Line 279). 

In short, we observed that the composite effect was larger in the monocular relative to CFS condition.

## Summary
With the above three glmm, we observed (1) composite face effect in the monocular condition and (2) it was larger than that in the CFS condition.

# RT: Lognormal
```{r}
# make the level names shorter
df_rt <- df_raw %>% 
  filter(isCorrect == 1)
```

## lmm for simple effects of Congruency 
```{r}
df_rt_sim <- model.matrix(~ Condition * Alignment / Congruency, df_rt) %>% 
  as_tibble() %>% 
  cbind(df_rt, .)
```
`df_rt_sim` was created to facilitate simplifying random effects later.

### The maximal model
```{r rt1 max}
# file_rt1_max <- file.path(folder_out, "RT1_lmm_max.RData")
# 
# # fit the max model
# if (!file.exists(file_rt1_max)) {
#   lmm_rt1_max <- lmer(log(RT) ~ Condition * Alignment / Congruency +
#                         (`Conditionmon-CFS` + `Alignmentali-mis` + 
#                            `Conditionmon-CFS:Alignmentali-mis` +
#                            `ConditionCFS:Alignmentali:Congruencycon-inc` + # interest
#                            `Conditionmon:Alignmentali:Congruencycon-inc` + # interest
#                            `ConditionCFS:Alignmentmis:Congruencycon-inc` +
#                            `Conditionmon:Alignmentmis:Congruencycon-inc`
#                          | Participant),
#                       data = df_rt_sim,
#                       control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
#                                             optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
#   )
# 
#   save(lmm_rt1_max, file = file_rt1_max)
# } else {
#   load(file_rt1_max)
# }
# 
# print(summary(lmm_rt1_max), corr = FALSE)
```

### The zero-correlation-parameter model
```{r rt1 zcp}
file_rt1_zcp <- file.path(folder_out, "RT1_lmm_zcp.RData")

# fit the zcp model
if (!file.exists(file_rt1_zcp)) {
  lmm_rt1_zcp <- lmer(log(RT) ~ Condition * Alignment / Congruency +
                        (`Conditionmon-CFS` + `Alignmentali-mis` + 
                           `Conditionmon-CFS:Alignmentali-mis` +
                           `ConditionCFS:Alignmentali:Congruencycon-inc` + # interest
                           `Conditionmon:Alignmentali:Congruencycon-inc` + # interest
                           `ConditionCFS:Alignmentmis:Congruencycon-inc` +
                           `Conditionmon:Alignmentmis:Congruencycon-inc`
                         || Participant),
                      data = df_rt_sim,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(lmm_rt1_zcp, file = file_rt1_zcp)
} else {
  load(file_rt1_zcp)
}

print(summary(lmm_rt1_zcp), corr = FALSE)
```

### The reduced model
```{r PCA analysis for rt1 zcp lmm }
summary(rePCA(lmm_rt1_zcp))
```

Random effect that explained less than 0.1% variances were removed from the zcp model, leading to the reduced model.

```{r  rt1 rdc}
file_rt1_rdc <- file.path(folder_out, "RT1_lmm_rdc.RData")

# fit the rdc model
if (!file.exists(file_rt1_rdc)) {
  lmm_rt1_rdc <- lmer(log(RT) ~ Condition * Alignment / Congruency +
                        (`Conditionmon-CFS` + # `Alignmentali-mis` + 
                           `Conditionmon-CFS:Alignmentali-mis` # +
                           # `ConditionCFS:Alignmentali:Congruencycon-inc` + # interest
                           # `Conditionmon:Alignmentali:Congruencycon-inc` + # interest
                           # `ConditionCFS:Alignmentmis:Congruencycon-inc` +
                           # `Conditionmon:Alignmentmis:Congruencycon-inc`
                         || Participant),
                      data = df_rt_sim,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(lmm_rt1_rdc, file = file_rt1_rdc)
} else {
  load(file_rt1_rdc)
}

print(summary(lmm_rt1_rdc), corr = FALSE)
```

### The extended model
```{r  rt1 etd}
file_rt1_etd <- file.path(folder_out, "RT1_lmm_etd.RData")

# fit the etd model
if (!file.exists(file_rt1_etd)) {
  lmm_rt1_etd <- lmer(log(RT) ~ Condition * Alignment / Congruency +
                        (`Conditionmon-CFS` + # `Alignmentali-mis` + 
                           `Conditionmon-CFS:Alignmentali-mis` # +
                           # `ConditionCFS:Alignmentali:Congruencycon-inc` + # interest
                           # `Conditionmon:Alignmentali:Congruencycon-inc` + # interest
                           # `ConditionCFS:Alignmentmis:Congruencycon-inc` +
                           # `Conditionmon:Alignmentmis:Congruencycon-inc`
                         | Participant),
                      data = df_rt_sim,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(lmm_rt1_etd, file = file_rt1_etd)
} else {
  load(file_rt1_etd)
}

print(summary(lmm_rt1_etd), corr = FALSE)
```


### The optimal model
```{r  comapre rt1 etd and rdc}
# compare the extended and reduced model
anova(lmm_rt1_etd, lmm_rt1_rdc, refit = FALSE)
```

```{r  the optimal model rt1}
lmm_rt1_opt <- lmm_rt1_rdc

print(summary(lmm_rt1_opt), corr = FALSE)
```
Results showed (the last row) that response times were shorter on congruent trials than incongruent trials for aligned faces in the monocular condition.


## lmm for composite effects in CFS and monocular conditions
```{r}
df_rt_cf <- model.matrix(~ Condition / Congruency * Alignment, df_rt) %>% 
  as_tibble() %>% 
  cbind(df_rt, .)
```
`df_rt_cf` was created to facilitate simplifying random effects later.

### The maximal model
```{r rt2 max}
# file_rt2_max <- file.path(folder_out, "RT2_lmm_max.RData")
# 
# # fit the max model
# if (!file.exists(file_rt2_max)) {
#   lmm_rt2_max <- lmer(log(RT) ~ Condition / Congruency * Alignment +
#                         (`Conditionmon-CFS` + `Alignmentali-mis` +
#                            `ConditionCFS:Congruencycon-inc` +
#                            `Conditionmon:Congruencycon-inc` +
#                            `Conditionmon-CFS:Alignmentali-mis` +
#                            `ConditionCFS:Congruencycon-inc:Alignmentali-mis` + # interest
#                            `Conditionmon:Congruencycon-inc:Alignmentali-mis`  # interest
#                          | Participant),
#                       data = df_rt_cf,
#                       control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
#                                             optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
#   )
# 
#   save(lmm_rt2_max, file = file_rt2_max)
# } else {
#   load(file_rt2_max)
# }
# 
# print(summary(lmm_rt2_max), corr = FALSE)
```

### The zero-correlation-parameter model
```{r rt2 zcp}
file_rt2_zcp <- file.path(folder_out, "RT2_lmm_zcp.RData")

# fit the zcp model
if (!file.exists(file_rt2_zcp)) {
  lmm_rt2_zcp <- lmer(log(RT) ~ Condition / Congruency * Alignment +
                        (`Conditionmon-CFS` + `Alignmentali-mis` +
                           `ConditionCFS:Congruencycon-inc` +
                           `Conditionmon:Congruencycon-inc` +
                           `Conditionmon-CFS:Alignmentali-mis` +
                           `ConditionCFS:Congruencycon-inc:Alignmentali-mis` + # interest
                           `Conditionmon:Congruencycon-inc:Alignmentali-mis`  # interest
                         || Participant),
                      data = df_rt_cf,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(lmm_rt2_zcp, file = file_rt2_zcp)
} else {
  load(file_rt2_zcp)
}

print(summary(lmm_rt2_zcp), corr = FALSE)
```

### The reduced model
```{r PCA analysis for rt2 zcp lmm }
summary(rePCA(lmm_rt2_zcp))
```

Random effect that explained less than 0.1% variances were removed from the zcp model, leading to the reduced model.
```{r  rt2 rdc}
file_rt2_rdc <- file.path(folder_out, "RT2_lmm_rdc.RData")

# fit the rdc model
if (!file.exists(file_rt2_rdc)) {
  lmm_rt2_rdc <- lmer(log(RT) ~ Condition / Congruency * Alignment +
                        (`Conditionmon-CFS` + # `Alignmentali-mis` +
                           # `ConditionCFS:Congruencycon-inc` +
                           # `Conditionmon:Congruencycon-inc` +
                           `Conditionmon-CFS:Alignmentali-mis` # +
                           # `ConditionCFS:Congruencycon-inc:Alignmentali-mis` + # interest
                           # `Conditionmon:Congruencycon-inc:Alignmentali-mis`  # interest
                         || Participant),
                      data = df_rt_cf,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )

  save(lmm_rt2_rdc, file = file_rt2_rdc)
} else {
  load(file_rt2_rdc)
}

print(summary(lmm_rt2_rdc), corr = FALSE)
```

### The extended model
```{r  rt2 etd}
file_rt2_etd <- file.path(folder_out, "RT2_lmm_etd.RData")

# fit the etd model
if (!file.exists(file_rt2_etd)) {
  lmm_rt2_etd <- lmer(log(RT) ~ Condition / Congruency * Alignment +
                        (`Conditionmon-CFS` + # `Alignmentali-mis` +
                           # `ConditionCFS:Congruencycon-inc` +
                           # `Conditionmon:Congruencycon-inc` +
                           `Conditionmon-CFS:Alignmentali-mis` # +
                           # `ConditionCFS:Congruencycon-inc:Alignmentali-mis` + # interest
                           # `Conditionmon:Congruencycon-inc:Alignmentali-mis`  # interest
                         | Participant),
                      data = df_rt_cf,
                      control = lmerControl(optimizer = "optimx", # calc.derivs = FALSE,
                                            optCtrl = list(method = "nlminb", starttests = FALSE, kkt = FALSE))
  )


  save(lmm_rt2_etd, file = file_rt2_etd)
} else {
  load(file_rt2_etd)
}

print(summary(lmm_rt2_etd), corr = FALSE)
```

### The optimal model
```{r  comapre rt2 etd and rdc}
# compare the extended and reduced model
anova(lmm_rt2_etd, lmm_rt2_rdc, refit = FALSE)
```

```{r  the optimal model rt2}
lmm_rt2_opt <- lmm_rt2_rdc

print(summary(lmm_rt2_opt), corr = FALSE)
```
Results showed that the interaction between Congruency and Alignment of response times was found in the CFS, but not the monocular condition. 

## lmm for comparing composite effects in CFS and monocular conditions
This analysis will not be repeated; it is the same as that used in `lmm_CF_emmeans.Rmd` file (around Line 430). 

In short, we failed to observe the evidence that the composite effects differed between monocular and CFS condition.

## Summary
With the above three glmm, we failed to observed the composite face effect in the either condition or the differences of composite face effect between the monocular and CFS condition.

# Session information {.unlisted .unnumbered}
```{r}
sessionInfo()
```

